Transcriptional dynamics of influenza virus infection in single cells

### 2017-09-01 paper/paper.tex
edits to first section of Methods

Infected cells will express viral as well as cellular mRNAs -- however the cell barcodes and UMIs cannot distinguish whether a cell was initially infected by one or multiple viral particles.
We therefore engineered an influenza virus (strain A/WSN/1933) that additionally carried \emph{viral barcodes} consisting of synonymous mutations near the 3' end of each transcript (Figure~\ref{fig:workflow}A).
Critically, these synonymous mutations did not greatly impact viral growth kinetics (Figure~\ref{fig:workflow}B).
We infected A549 human lung carcinoma cells with an equal mix of the wild-type and synonymously barcoded viruses.
Cells infected by a single virion will exclusively express mRNAs from either wild-type or synonymously barcoded virus, whereas cells that are co-infected with multiple virions will often express mRNAs from both the wild-type and synonymously barcoded viruses (Figure~\ref{fig:workflow}C).
\begin{figure}
\centerline{\includegraphics[width=0.8\linewidth]{figures/Workflow/workflow.pdf}}
\caption{\label{fig:workflow} Experimental design.
{\bf (A)}  
We engineered a virus that carried two synonymous mutations near the 3' end of each mRNA.
At top are the specific mutations for PB2.
At bottom are locations of the synonymous mutations relative to the typical distribution of read depth for our 3'-end sequencing.
{\bf (B)} 
The wild-type and synonymously barcoded viruses transcribe their genes with similar kinetics. 
Shown is the abundance of the viral hemagglutinin (HA) transcript relative to the cellular housekeeping gene L32 as assessed by qPCR in A549 cells infected at an MOI of 0.5 (as determined on MDCK-SIAT1 cells).
Error bars $\pm$ S.D., n=3.
{\bf (C)}  
A549 cells were infected with an equal mixture of mutant and wild-type virus. 
Immediately prior to collection, cells were physically separated into droplets and cDNA libraries were generated containing the indicated barcodes. 
The libraries were deep sequenced, and the data processed to create a matrix that gives the number of molecules of each transcript observed in each cell.
Infected cells were further annotated by whether their viral mRNAs derived from wild-type virus, synonymously barcoded virus, or both.
\jdbcomment{Make the matrix more in line with the data that we actually see. Make the counts of the different host genes reasonably similar between cells. We think most host genes have pretty similar expression across cells, but your example shows all having at least 2-fold difference and some much larger. So adjust the numbers to be more similar among cells. Also, in A and the matrix in C, just write out ``wildtype'', ``synonymous'', and ``not detected'' (or ``none``) -- there is plenty of space, so no need to use acronyms.}
}
\figdata{Sequences of wild-type and barcoded viruses are in \texttt{viralsequences.fasta}.}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=0.9\linewidth]{figures/p_coinfection.pdf}}
\caption{\label{fig:coexpression}
The abundance of each viral transcript in cells that are co-infected with the two viral variants and have $>$5\% of their mRNA derived from virus.
The bars show the logarithms of the numbers of each viral mRNA detected, and are colored in linear proportion to the fraction of that mRNAs derived from wild-type or synonymously barcoded virus.
}
\figsupp[Co-infected cells express roughly equal amounts of a gene from each infecting viral variant. \label{figsupp:flowcyto}]{
{\bf(A)} Cells were co-infected with a mix of wild-type virus and virus in which the HA gene was replaced by GFP flanked by the terminal regions of the HA gene segment. 
At 10 hours post-infection, cells were analyzed by flow cytometry for HA and eGFP expression.
{\bf(B)} The expression of HA and GFP are correlated in co-infected cells. 
Shown are the quantile-normalized HA and eGFP signals for double-positive cells. 
Cells are colored by density, using a Gaussian kernel density estimate. 
{\bf(C)},{\bf(D)},{\bf(E)} Gating controls, single infection with eGFP virus, single infection with wild-type virus, and uninfected cells, respectively. 
}{\includegraphics[width=\linewidth]{figures/Pseudovirus_flow_cytometry/coinfFlow_D02.png}}
\figdata{The raw data plotted in this figure are in \texttt{p\_co-infection.csv}.}
\end{figure}

@@

\subsection{Tissue culture and viral rescue}
Cell lines used in this study were: the human lung epithelial carcinoma line, A549, obtained from ATCC, the canine kidney cell line overexpressing human SIAT1, MDCK-SIAT1, obtained from Sigma Aldrich, and the human embryonic kidney cell line, 293t, obtained from ATCC. 
All cells were maintained in D10 (DMEM supplemented with 10\% heat-inactivated FBS, 2 mM L-glutamine, 100 U of penicillin/ml, and 100 \si{\micro}g of streptomycin/ml) at 37 \si{\degreeCelsius} at 5 \% CO\textsubscript{2}.
We generated viral stocks using a previously-defined reverse-genetics approach~\citep{Neumann:1999ws}. 
In brief, to generate wild-type virus we transfected bidirectional plasmids encoding the eight segments of A/WSN/1933 (kindly provided by Robert Webster of St. Jude Children's Research Hospital) into a coculture of 293t and MDCK-SIAT1 cells seeded at a ratio of 8:1. 
At 24 hours post-transfection, we changed media from D10 to influenza growth media (IGM, Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/ml, 100  \si{\micro}g of streptomycin/ml, and 100 \si{\micro}g of calcium chloride/ml).
At 48 hours post-transfection we harvested the virus-containing supernatant, removed cellular material by centrifugation at 300 x g's for 4 minutes, and stored aliquots at -80 \si{\degreeCelsius }. 
We then determined the TCID50 of these stocks on MDCK-SIAT1 cells.
Viral aliquots were then used to initiate an infection at an MOI of 0.01 on MDCK-SIAT1 cells in order to expand viral stocks and provide partial purging of defective viruses.
Viral supernatant was harvested at 36 hours post-infection, prior to the onset of noticeable cytopathic effect. 
Aliquots were again frozen at -80  \si{\degreeCelsius }, titered, and used in our single-cell experiments.
Synonymously barcoded virus was generated identically to wild-type viral stocks, using plasmids encoding the synonymously-encoded viral segments described in (supplemental data).
Mutant virus containing a gene encoding eGFP in place of the HA gene segment was generated using previously-described methods for growing HA-deficient virus, utilizing MDCK-SIAT1 cells expressing HA from a constituitive CMV promoter~\citep{Doud:2016gm}.
A unidirectional vector for generating viral vRNA encoding eGFP flanked by HA packaging sequences was generated from a previously described vector, pHH21~\citep{Neumann:1999ws}. 
In order to obtain sufficient titers, HA-eGFP virus was expanded for 44 hours after initiating infection at an MOI of 0.01 rather than 36 hours.

\subsection{qPCR measurements of infection kinetics}
For qPCR measurements of viral transcriptional kinetics and interferon induction, A549 cells were seeded at 3 x 10\textsuperscript{5} cells / well in a 6-well tissue culture plate in D10 the day prior to infection. 
On the day of infection, a single well was trypsonized and enumerated, providing a measurement for MOI calculation.
Immediately before infecting, D10 was replaced with IGM and the indicated treatment applied.
For cells incubated with cycloheximide, the compound was added to a final concentration of 50  \si{\micro}g/ml at the time of infection -- previously confirmed to be sufficient to halt viral protein production~\citep{Killip:2014kz}.
RNA was purified using the QIAGEN RNeasy plus mini kit following manufacturer's suggestions, and purity and concentration was measured using UV absorbance prior to cDNA synthesis. 
cDNA was synthesized using an oligoDT primer and the SuperScript\textsuperscript{TM} III first-strand synthesis supermix from ThermoFisher using manufacturer's directions. 
Transcript abundance was measured using SYBR\textsuperscript{TM}  green PCR master mix, using a combined anneal/extension step of 60 \si{\degreeCelsius } for one minute with the primers listed in (Supplemental table).
Biological triplicates were performed for all samples.
For measurements of viral genomic HA content, vRNA was harvested from 80 \si{\micro}l of viral supernatant by the addition of 600 \si{\micro}l of RLT plus before proceeding with the standard QIAGEN Rneasy plus miniprep kit protocol, and cDNA was generated using  SuperScript\textsuperscript{TM} III first-strand synthesis supermix with primers specific for vRNA using the manufacturer's protocol for random-hexamer priming~\citep{Xue:2017dl,Hoffmann:2001vj}.
qPCR was then performed as for mRNA measurements.
In order to obtain molecular resolution on our measurements of HA vRNA, a standard curve was generated from three independent dilutions of the HA-encoding bidirectional plasmid used in our viral rescue. 
All vRNA values represent three independent RNA extractions with two replicate qPCR measurements.

@@

\subsection{Generation of single cell infection data}
All samples were processed using the 10x Chromium Single Cell 3' platform, both in initial, tagged, cDNA generation and in final library preparation.
All time points, save for the 8 hour repeat, were performed on the same day as described below.
A549 cells were seeded as described above in a 6-well plate, with two wells per time point. 
A single well of cells was trypsonized and counted prior to initiation of the experiment for the purposes of calculating MOI.
Wild-type and synonymously barcoded virus were mixed to an estimated ratio of 1:1 based on prior, exploratory, single-cell analyses (data not shown). 
At the initiation of our experiment, all cells were changed from D10 to IGM, regardless of the desired time point.
Cells were then infected with an intended, flow cytometry calculated, MOI of 0.3, in order of time point -- 10 hour, then 8 hour, then 6 hour time points, spaced two hours apart.
This MOI was calculated in the absence of an additional washing step, and so potentially represents an overestimate of our final infectious dose.
One hour post infection, media was changed to fresh IGM.
All cells were then harvested for single-cell analysis concurrently -- ensuring all had spent equivalent time in changed media and normalizing potential batch effects across samples.
For our replicate 8 hour time point, cells were infected as above, save for no wash step was included, the flow cytometry informed MOI was instead 0.1, and the sample was prepared on a different day.
After harvest, cells were counted using disposable hemocytometers and diluted to equivalent concentrations with an intended capture of 3000 cells/sample following the guidelines provided by 10x genomics.
All subsequent steps followed 10x genomics guidelines through library preparation.
Samples were sequenced on an Illumina HiSeq.

>>>>

Infected cells will express viral as well as cellular mRNAs -- however the cell barcodes and UMIs cannot distinguish whether a cell was initially infected by one or multiple viral particles.
We therefore engineered an influenza virus (strain A/WSN/1933) that additionally carried \emph{viral barcodes} consisting of synonymous mutations near the 3' end of each transcript (Figure~\ref{fig:workflow}A).
Critically, these synonymous mutations did not greatly impact viral growth kinetics (Figure~\ref{fig:workflow}B).
We infected A549 human lung carcinoma cells with an equal mix of the wild-type and synonymously barcoded viruses.
Cells infected by a single virion will exclusively express mRNAs from either wild-type or synonymously barcoded virus, whereas cells that are co-infected with multiple virions will often express mRNAs from both the wild-type and synonymously barcoded viruses (Figure~\ref{fig:workflow}C).
\begin{figure}
\centerline{\includegraphics[width=0.8\linewidth]{figures/Workflow/workflow.pdf}}
\caption{\label{fig:workflow} Experimental design.
{\bf (A)}  
We engineered a virus that carried two synonymous mutations near the 3' end of each mRNA.
At top are the specific mutations for PB2.
At bottom are locations of the synonymous mutations relative to the typical distribution of read depth for our 3'-end sequencing.
{\bf (B)} 
The wild-type and synonymously barcoded viruses transcribe their genes with similar kinetics. 
Shown is the abundance of the viral hemagglutinin (HA) transcript relative to the cellular housekeeping gene L32 as assessed by qPCR in A549 cells infected at an MOI of 0.5 (as determined on MDCK-SIAT1 cells).
Error bars $\pm$ S.D., n=3.
{\bf (C)}  
A549 cells were infected with an equal mixture of mutant and wild-type virus. 
Immediately prior to collection, cells were physically separated into droplets and cDNA libraries were generated containing the indicated barcodes. 
The libraries were deep sequenced, and the data processed to create a matrix that gives the number of molecules of each transcript observed in each cell.
Infected cells were further annotated by whether their viral mRNAs derived from wild-type virus, synonymously barcoded virus, or both.
\jdbcomment{Make the matrix more in line with the data that we actually see. Make the counts of the different host genes reasonably similar between cells. We think most host genes have pretty similar expression across cells, but your example shows all having at least 2-fold difference and some much larger. So adjust the numbers to be more similar among cells. Also, in A and the matrix in C, just write out ``wildtype'', ``synonymous'', and ``not detected'' (or ``none``) -- there is plenty of space, so no need to use acronyms.}
}
\figdata{Sequences of wild-type and barcoded viruses are in \texttt{viralsequences.fasta}.\label{figdata:viralsequences}}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=0.9\linewidth]{figures/p_coinfection.pdf}}
\caption{\label{fig:coexpression}
The abundance of each viral transcript in cells that are co-infected with the two viral variants and have $>$5\% of their mRNA derived from virus.
The bars show the logarithms of the numbers of each viral mRNA detected, and are colored in linear proportion to the fraction of that mRNAs derived from wild-type or synonymously barcoded virus.
}
\figsupp[Co-infected cells express roughly equal amounts of a gene from each infecting viral variant. \label{figsupp:flowcyto}]{
{\bf(A)} Cells were co-infected with a mix of wild-type virus and virus in which the HA gene was replaced by GFP flanked by the terminal regions of the HA gene segment. 
At 10 hours post-infection, cells were analyzed by flow cytometry for HA and eGFP expression.
{\bf(B)} The expression of HA and GFP are correlated in co-infected cells. 
Shown are the quantile-normalized HA and eGFP signals for double-positive cells. 
Cells are colored by density, using a Gaussian kernel density estimate. 
{\bf(C)},{\bf(D)},{\bf(E)} Gating controls, single infection with eGFP virus, single infection with wild-type virus, and uninfected cells, respectively. 
}{\includegraphics[width=\linewidth]{figures/Pseudovirus_flow_cytometry/coinfFlow_D02.png}}
\figdata{The raw data plotted in this figure are in \texttt{p\_co-infection.csv}.}
\figdata{The sequence of the HA viral RNA carrying the GFP gene is in \texttt{HAflank-eGFP.fasta}.\label{figdata:HAflank}}
\end{figure}

@@

\subsection{Cell lines and viruses}
The following cell lines were used in this study: the human lung epithelial carcinoma line A549 (ATCC CCL-185), the MDCK-SIAT1 variant of the Madin Darby canine kidney cell line overexpressing human SIAT1 (Sigma-Aldrich 05071502), and the human embryonic kidney cell line 293T (ATCC CRL-3216). 
All cells were maintained in D10 media (DMEM supplemented with 10\% heat-inactivated fetal bovine serum, 2 mM L-glutamine, 100 U of penicillin/ml, and 100 \si{\micro}g of streptomycin/ml) at 37 \si{\degreeCelsius} at 5 \% CO\textsubscript{2}.

Wildtype A/WSN/1933 (H1N1) influenza virus was generated by reverse genetics using the plasmids pHW181-PB2, pHW182-PB1, pHW183-PA, pHW184-HA, pHW185-NP, pHW186-NA, pHW187-M, and pHW188-NS~\citep{hoffmann2000dna}.
The sequences of the viral RNAs encoded in these plasmids are in Figure~\ref{fig:workflow}-source~data~\ref{figdata:viralsequences}.
Reverse-genetics plasmids encoding the synonymously barcoded WSN virus were created by using site-directed mutagenesis to introduce two synonymous mutations near the 3' end of the mRNA for each viral gene.
The synonymous mutations were chosen to introduce nucleotides that are common in naturally occurring H1N1 viruses to avoid unwanted attenuation.
The sequences of the synonymously barcoded viral RNAs are in Figure~\ref{fig:workflow}-source~data~\ref{figdata:viralsequences}
To generate viruses from these plasmids, we transfected an equimolar mix of all eight plasmids into cocultures of 293T and MDCK-SIAT1 cells seeded at a ratio of 8:1. 
At 24 hours post-transfection, we changed media from D10 to influenza growth media (Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/ml, 100  \si{\micro}g of streptomycin/ml, and 100 \si{\micro}g of calcium chloride/ml).
At 48 hours post-transfection we harvested the virus-containing supernatant, pelletted cellular material by centrifugation at 300 x g's for 4 minutes, and stored aliquots at of the clarified viral supernatant -80 \si{\degreeCelsius }. 
We then titered thawed aliquots of viral by TCID50 on MDCK-SIAT1 cells, computing titers via the formula of \citet{reed1938simple}.
To generate our ``high-purity'' stocks of viruses for the single-cell sequencing experiments, we then infected MDCK-SIAT1 cells at an MOI of 0.01, and let the virus replicate for 36 hours prior to harvesting aliquots that were again clarified by low-speed centrifugation, aliquoted, stored at -80  \si{\degreeCelsius }, and titered by TCID50.

The high-MOI passage (high-defective particle) stock used in Figure~\ref{fig:viruspopulations} was generated by \jdbcomment{Alistair, add this.}

For the experiments in Figure~\ref{fig:coexpression}-Figure~supplement~\ref{figsupp:flowcyto}, we created a virus that carried an HA gene segment in which the carried GFP in place of most of the HA coding sequence by following a scheme first described by \citet{marsh2007specific}.
Briefly, we created a plasmid encoding a viral RNA with GFP in place of the HA coding sequence in the context of the pHH21~\citep{Neumann:1999ws} reverse-genetics plasmid, removing potential start codons upstream of the GFP (see Figure~\ref{fig:coexpression}-source~data~\ref{figdata:HAflank} for the sequence of the viral RNA)
We then generated GFP-carrying virus by reverse-genetics in previously described cells constitutively expressing HA~\citep{Doud:2016gm}.
A unidirectional vector for generating viral vRNA encoding eGFP flanked by HA packaging sequences was generated from a previously described vector, pHH21~\citep{Neumann:1999ws}. 
To obtain sufficient titers, this HA-eGFP virus was expanded for 44 than than 36 hours after initiating infection at an MOI of 0.01.

\subsection{qPCR measurements of infection kinetics}
For qPCR measurements of viral transcriptional kinetics and interferon induction, A549 cells were seeded at 3 x 10\textsuperscript{5} cells / well in a 6-well tissue culture plate in D10 the day prior to infection. 
On the day of infection, a single well was trypsonized and enumerated, providing a measurement for MOI calculation.
Immediately before infecting, D10 was replaced with influenza growth media and the indicated treatment applied.
For cells incubated with cycloheximide, the compound was added to a final concentration of 50  \si{\micro}g/ml at the time of infection -- previously confirmed to be sufficient to halt viral protein production~\citep{Killip:2014kz}.
RNA was purified using the QIAGEN RNeasy plus mini kit following manufacturer's suggestions, and purity and concentration was measured using UV absorbance prior to cDNA synthesis. 
cDNA was synthesized using an oligoDT primer and the SuperScript\textsuperscript{TM} III first-strand synthesis supermix from ThermoFisher using manufacturer's directions. 
Transcript abundance was measured using SYBR\textsuperscript{TM}  green PCR master mix, using a combined anneal/extension step of 60 \si{\degreeCelsius } for one minute with the primers listed in (Supplemental table).
Biological triplicates were performed for all samples.
For measurements of viral genomic HA content, vRNA was harvested from 80 \si{\micro}l of viral supernatant by the addition of 600 \si{\micro}l of RLT plus before proceeding with the standard QIAGEN Rneasy plus miniprep kit protocol, and cDNA was generated using  SuperScript\textsuperscript{TM} III first-strand synthesis supermix with primers specific for vRNA using the manufacturer's protocol for random-hexamer priming~\citep{Xue:2017dl,Hoffmann:2001vj}.
qPCR was then performed as for mRNA measurements.
In order to obtain molecular resolution on our measurements of HA vRNA, a standard curve was generated from three independent dilutions of the HA-encoding bidirectional plasmid used in our viral rescue. 
All vRNA values represent three independent RNA extractions with two replicate qPCR measurements.

@@

\subsection{Generation of single cell infection data}
All samples were processed using the 10x Chromium Single Cell 3' platform, both in initial, tagged, cDNA generation and in final library preparation.
All time points, save for the 8 hour repeat, were performed on the same day as described below.
A549 cells were seeded as described above in a 6-well plate, with two wells per time point. 
A single well of cells was trypsonized and counted prior to initiation of the experiment for the purposes of calculating MOI.
Wild-type and synonymously barcoded virus were mixed to an estimated ratio of 1:1 based on prior, exploratory, single-cell analyses (data not shown). 
At the initiation of our experiment, all cells were changed from D10 to influenza growth media, regardless of the desired time point.
Cells were then infected with an intended, flow cytometry calculated, MOI of 0.3, in order of time point -- 10 hour, then 8 hour, then 6 hour time points, spaced two hours apart.
This MOI was calculated in the absence of an additional washing step, and so potentially represents an overestimate of our final infectious dose.
One hour post infection, media was changed to fresh influenza growth media.
All cells were then harvested for single-cell analysis concurrently -- ensuring all had spent equivalent time in changed media and normalizing potential batch effects across samples.
For our replicate 8 hour time point, cells were infected as above, save for no wash step was included, the flow cytometry informed MOI was instead 0.1, and the sample was prepared on a different day.
After harvest, cells were counted using disposable hemocytometers and diluted to equivalent concentrations with an intended capture of 3000 cells/sample following the guidelines provided by 10x genomics.
All subsequent steps followed 10x genomics guidelines through library preparation.
Samples were sequenced on an Illumina HiSeq.
